Plug-and-play dual-tree algorithm runtime analysis

### 2014-11-13 bounds-files/imbalance.tex
Refactor some text; add the figures; fix the citations.

A better understanding of the behavior of cover trees can be acheived partially
by introducing a measure of tree imbalance.  Numerous measures of tree imbalance
have already been established, but we aim to capture a different measure of
imbalance that utilizes the leveled structure of the cover tree.

We already know each node in a cover tree is indexed with an integer level (or
scale).  In the explicit representation of the cover tree, each non-leaf node
has children at a lower level.  But these children need not be strictly one
level lower; see Figure \ref{fig:imbalance}.  In Figure
\ref{fig:imbalance-good}, each cover tree node has children that are strictly
one level lower; we will refer to this as a {\em perfectly balanced cover tree}.
Figure \ref{fig:imbalance-bad}, on the other hand, has many children with far
lower levels than their parents.  We will refer to this as an {\em imbalanced
cover tree}.

\begin{figure}
\caption{Balanced and imbalanced cover trees.}
\label{fig:imbalance}
\end{figure}

An imbalanced cover tree can---and often does---happen in practice.  Consider a
dataset with a single outlier which is very far away from all of the other
points.  Figure \ref{fig:outlier} shows what happens in this situation: the root
node has two children; one of these children has only the outlier as a
descendant, and the other child has the rest of the points in the dataset as a
descendant.  In fact, it is easy to find datasets with multiple outliers which
give rise to a chain-like structure at the top of the tree: see Figure
\ref{fig:outliers} for an illustration\footnote{As a side note, this behavior is
not limited to cover trees, and can happen to mean-split $kd$-trees too,
especially in higher dimensions.}.

\begin{figure}
\caption{Single-outlier cover tree.}
\label{fig:outlier}
\end{figure}

\begin{figure}
\caption{Multiple-outlier cover tree.}
\label{fig:outliers}
\end{figure}

A tree that has this chain-like structure all the way down is going to perform
abysmally; motivated by this observation, we define a measure of tree imbalance.

>>>>

It is well-known that imbalance in trees leads to degradation in performance;
for instance, a $kd$-tree node with every descendant except one in its left
child is effectively useless.  A $kd$-tree full of nodes like this will perform
abysmally for nearest neighbor search, and it is not hard to generate a dataset
that will cause a $kd$-tree of this sort.

This sort of imbalance applies to all types of trees, not just $kd$-trees.  In
our situation, we are interested in a better understanding of this imbalance for
cover trees, and thus endeavor to introduce a more formal measure of imbalance
which is correlated with tree performance.  Numerous measures of tree
imbalance have already been established; one example is that proposed by
\citet{colless1982review}, and another is Sackin's index \citep{sackin1972good},
but we aim to capture a different measure of imbalance that utilizes the leveled
structure of the cover tree.

We already know each node in a cover tree is indexed with an integer level (or
scale).  In the explicit representation of the cover tree, each non-leaf node
has children at a lower level.  But these children need not be strictly one
level lower; see Figure \ref{fig:imbalance}.  In Figure
\ref{fig:imbalance-good}, each cover tree node has children that are strictly
one level lower; we will refer to this as a {\em perfectly balanced cover tree}.
Figure \ref{fig:imbalance-bad}, on the other hand, contains the node
$\mathscr{N}_m$ which has two children with scale two less than $s_m$.  We will
refer to this as an {\em imbalanced cover tree}.  Note that in our definition,
the balance of a cover tree has nothing to do with differing number of
descendants in each child branch but instead only missing levels.

\begin{figure}
\begin{subfigure}[b]{0.585\textwidth}
  \begin{center}
    \input{figures/balanced-cover-tree.tex}
  \end{center}
  \caption{Balanced cover tree.}
  \label{fig:imbalance-good}
\end{subfigure}
\begin{subfigure}[b]{0.415\textwidth}
  \begin{center}
    \input{figures/imbalanced-cover-tree.tex}
  \end{center}
  \caption{Imbalanced cover tree.}
  \label{fig:imbalance-bad}
\end{subfigure}
\caption{Balanced and imbalanced cover trees.}
\label{fig:imbalance}
\end{figure}

An imbalanced cover tree can---and often does---happen in practice, and the
imbalance may be far worse than the simple graphs of Figure \ref{fig:imbalance}.
Consider a dataset with a single outlier which is very far away from all of the
other points\footnote{Note also that for an outlier sufficiently far away, the
expansion constant is $N - 1$.}.  Figure \ref{fig:outlier} shows what happens in
this situation: the root node has two children; one of these children has only
the outlier as a descendant, and the other child has the rest of the points in
the dataset as a descendant.  In fact, it is easy to find datasets with multiple
outliers which give rise to a chain-like structure at the top of the tree: see
Figure \ref{fig:outliers} for an illustration\footnote{As a side note, this
behavior is not limited to cover trees, and can happen to mean-split $kd$-trees
too, especially in higher dimensions.}.

\begin{figure}
\begin{center}
  \input{figures/single-outlier-tree.tex}
\end{center}
\caption{Single-outlier cover tree.}
\label{fig:outlier}
\end{figure}

\begin{figure}
\begin{center}
  \input{figures/multiple-outlier-tree.tex}
\end{center}
\caption{A badly-drawn multiple-outlier cover tree.}
\label{fig:outliers}
\end{figure}

A tree that has this chain-like structure all the way down is going to perform
horrendously; motivated by this observation, we define a measure of tree
imbalance.

### 2014-11-13 bounds-files/sec01-supporting-lemmas.tex
Refactor some text; add the figures; fix the citations.

For simplicity, the algorithms considered in this paper will be presented in a
tree-independent context, as in \cite{curtin2013tree}, but the only type of tree
we will consider is the cover tree \cite{langford2006}, and the only type of
traversal we will consider is the cover tree pruning dual-tree traversal, which
we will describe below.

As we will be making heavy use of trees, we must establish notation (taken from
\cite{curtin2013tree}).  The notation we will be using is defined in Table
\ref{tab:notation}.

@@

The cover tree is a leveled hierarchical data structure originally proposed for
the task of nearest neighbor search by Beygelzimer
et~al.~\cite{langford2006}.  The original description of the tree describes both
an {\it implicit} and {\it explicit} representation, where the explicit
representation is the actual structure that would be implemented in practice.
Thus, we will consider only the explicit representation in this paper.

@@

A batch construction algorithm is given in \cite{langford2006}.  In addition,
the cover tree has a number of other useful properties and lemmas based on the
expansion constant \cite{karger2002finding}; we restate the definition below.

@@

The expansion constant is used heavily in the cover tree literature.  It is
related to a notion of instrinic dimensionality, and in many scenarios, $c$ is
independent of the number of points in the dataset \cite{langford2006,
karger2002finding}.

@@

@@

\begin{equation}
|B_S(p, \rho \delta)| \le \frac{|B_S(p_i, 2 \rho \delta)|}{|B_S(p_i, \delta
/ 2)|} \le c^{2 + \log_2 \rho}.
\end{equation}\end{proof}

>>>>

For simplicity, the algorithms considered in this paper will be presented in a
tree-independent context, as in \citet{curtin2013tree}, but the only type of
tree we will consider is the cover tree \citep{langford2006}, and the only type
of traversal we will consider is the cover tree pruning dual-tree traversal,
which we will describe below.

As we will be making heavy use of trees, we must establish notation \citep[taken
from][]{curtin2013tree}.  The notation we will be using is defined in Table
\ref{tab:notation}.

@@

The cover tree is a leveled hierarchical data structure originally proposed for
the task of nearest neighbor search.  The original description of the tree
describes both an {\it implicit} and {\it explicit} representation, where the
explicit representation is the actual structure that would be implemented in
practice.  Thus, we will consider only the explicit representation in this
paper.

@@

A batch construction algorithm is given by \citet{langford2006}, called
\texttt{Construct}; but the details of construction are not relevant to our
work, so we only mention this as a reference.  More relevant to our work is the
number of useful and unique theoretical properties the cover tree possesses,
based on the expansion constant \citep{karger2002finding}; we restate the
definition below.

@@

The expansion constant is used heavily in the cover tree literature.  It is
related to a notion of instrinic dimensionality, and previous work has shown
that there are many scenarios where $c$ is independent of the number of points
in the dataset.

@@

Lastly, we introduce a convenience lemma of our own which is a generalization of
the packing arguments used by \citet{langford2006}.  This is a more flexible
version of their argument.

@@

\begin{equation}
|B_S(p, \rho \delta)| \le \frac{|B_S(p_i, 2 \rho \delta)|}{|B_S(p_i, \delta
/ 2)|} \le c^{2 + \log_2 \rho}.
\end{equation}
\end{proof}

### 2014-11-13 bounds.tex
Refactor some text; add the figures; fix the citations.

\documentclass{article} % For LaTeX2e
\usepackage{hyperref}
\hypersetup{colorlinks,
            urlcolor=blue,
            citecolor=green,
            linkcolor=red}
\usepackage{url}
%\documentstyle[nips13submit_09,times,art10]{article} % For LaTeX 2.09
\usepackage{mathrsfs}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{color}

\usepackage{pstricks}
\usepackage{tikz}
%\usepackage{subcaption}

\title{Plug-and-play dual-tree algorithm runtime analysis}
\author{Ryan R. Curtin, Dongryeol Lee, William B. March, Parikshit Ram}
\date{\today}

%\author{
%David S.~Hippocampus\thanks{ Use footnote for providing further information
%about author (webpage, alternative address)---\emph{not} for acknowledging
%funding agencies.} \\
%Department of Computer Science\\
%Cranberry-Lemon University\\
%Pittsburgh, PA 15213 \\
%\texttt{hippo@cs.cranberry-lemon.edu} \\
%\And
%Coauthor \\
%Affiliation \\
%Address \\
%\texttt{email} \\
%\AND
%Coauthor \\
%Affiliation \\
%Address \\
%\texttt{email} \\
%\And
%Coauthor \\
%Affiliation \\
%Address \\
%\texttt{email} \\
%\And
%Coauthor \\
%Affiliation \\
%Address \\
%\texttt{email} \\
%(if needed)\\
%}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

@@

\newtheorem{defn}{Definition}
\newtheorem{thm}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{cor}{Corollary}

@@

@@

A surprising number of machine learning algorithms have computational
bottlenecks that can be expressed as pairwise statistical problems.
Nearest neighbor search is one such problem, naively requiring $O(N)$ time
to answer in single query in a reference set of size $N$; answering $O(N)$
queries subsequently requires prohibitive $O(N^2)$ time. Kernel density
estimation is also a pairwise statistical problem, naively requiring $O(N^2)$
time to answer $O(N)$ queries. The reference set is typically indexed with
spatial data structures to accelerate this type of computation
\cite{finkel1974quad, langford2006}; these result in $O(\log N)$ runtime per
query under favorable conditions.

Gray and Moore \cite{nbody} generalized the fast multipole method from
computational physics to obtain dual-tree algorithms.  These are extremely
useful when there are large query sets, not just a few query points.  Instead of
building a tree on the reference set and searching with each query point
separately, Gray and Moore suggest also building a query tree and traversing
both the query and reference trees simultaneously (a {\it dual-tree traversal},
from which the class of algorithms takes its name).

Dual-tree algorithms can be easily understood through the recent framework of
Curtin et~al. \cite{curtin2013tree}: two trees (a query tree and a reference
tree) are traversed by a {\it pruning dual-tree traversal}.  This traversal
visits combinations of nodes from the trees in some sequence (each combination
consisting of a query node and a reference node), calling a problem-specific
\texttt{Score()} function to determine if the node combination can be pruned.
If not, then a problem-specific \texttt{BaseCase()} function is called for each
combination of points held in the query node and reference node.  This has
significant similarity to the more common single-tree branch-and-bound
algorithms, except that the algroithm must recurse into child nodes of {\it
both} the query tree and reference tree.

There exist numerous dual-tree algorithms for problems as diverse as kernel
density estimation \cite{gray2003nonparametric}, mean shift \cite{wang2007fast},
minimum spanning tree calculation \cite{march2010euclidean}, $n$-point
correlation function estimation \cite{march2012fast}, max-kernel search
\cite{curtin2013fast}, particle smoothing \cite{klaas2006fast}, variational
inference \cite{amizadeh2012variational}, and range search \cite{nbody}, to name
a few.

Some of these algorithms are derived using the cover tree \cite{langford2006}, a
data structure with beneficial theoretical qualities.  Dual-tree all
nearest-neighbor search and approximate kernel density estimation with cover
trees have $O(N)$ runtime guarantees for $O(N)$ queries \cite{ram2009}; minimum
spanning tree calculation scales as $O(N \log N)$ \cite{march2010euclidean}.
Other problems have similar worst-case guarantees \cite{curtin2014dual,
march2013multi}.

In this work we combine the generalization of Curtin et~al.
\cite{curtin2013tree} with the theoretical results of Beygelzimer et~al.
\cite{langford2006} (and others) in order to develop a worst-case runtime bound
for any dual-tree algorithm when the cover tree is used.

Section \ref{sec:preliminaries} lays out the required background, notation, and
introduces the cover tree and its associated theoretical properties.  Readers
familiar with the cover tree literature and dual-tree algorithms (especially
\cite{curtin2013tree}) may find that section to be review.  Following that, we
introduce an intuitive measure of cover tree imbalance in Section
\ref{sec:imbalance}.  This measure of imbalance is then used to prove the main
result of the paper in Section \ref{sec:bound}, which is a worst-case runtime
bound for generalized dual-tree algorithms.  We apply this result to three
specific problems: nearest neighbor search (Section \ref{sec:nns}), approximate
kernel density estimation (Section \ref{sec:akde}), and range search / range
count (Section \ref{sec:rs}), showing linear runtime bounds for each of those
algorithms.  Each of these bounds is an improvement on the state-of-the-art, and
in the case of range search, is the first such bound.  We then investigate the
empirical performance of cover trees for these algorithms, with a focus on
understanding the dataset-dependent constants that our bounds use.

@@

\bibliographystyle{plain}
\bibliography{bounds}

>>>>

\documentclass[twoside,11pt]{article} % For LaTeX2e
\usepackage{jmlr2e}
\usepackage{hyperref}
\hypersetup{colorlinks,
            urlcolor=blue,
            citecolor=green,
            linkcolor=red}
\usepackage{url}
%\documentstyle[nips13submit_09,times,art10]{article} % For LaTeX 2.09
\usepackage{mathrsfs}
%\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amsfonts}
%\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{color}

\usepackage{pstricks}
\usepackage{tikz}
\usepackage{subcaption}

@@

%\newtheorem{defn}{Definition}
%\newtheorem{thm}{Theorem}
%\newtheorem{lemma}{Lemma}
%\newtheorem{cor}{Corollary}
\newtheorem{defn}{Definition}
\newtheorem{thm}{Theorem}
\newtheorem{cor}{Corollary}

@@

\title{Plug-and-play dual-tree algorithm runtime analysis}

\author{\name Ryan R. Curtin \email ryan@ratml.org\\
  \addr School of Computational Science and Engineering\\
  Georgia Institute of Technology\\
  Atlanta, GA 30332-0250, USA
  \AND
  \name Dongryeol Lee \email drselee@gmail.com\\
  \addr GE Research\\
  Schenedectadadaty, NY\\
  \AND
  \name William B. March \email march@ices.utexas.edu\\
  \addr Austin, TX, USA\\
  \AND
  \name Parikshit Ram \email p.ram@gatech.edu\\
  \addr Skytree
}

\editor{Unknown}

@@

A surprising number of machine learning algorithms have computational
bottlenecks that can be expressed as pairwise statistical problems.
Nearest neighbor search is one such problem, naively requiring $O(N)$ time
to answer in single query in a reference set of size $N$; answering $O(N)$
queries subsequently requires prohibitive $O(N^2)$ time. Kernel density
estimation is also a pairwise statistical problem, naively requiring $O(N^2)$
time to answer $O(N)$ queries. The reference set is typically indexed with
spatial data structures to accelerate this type of computation
\citep{finkel1974quad, langford2006}; these result in $O(\log N)$ runtime per
query under favorable conditions.

\citet{nbody} generalized the fast multipole method from
computational physics to obtain dual-tree algorithms.  These are extremely
useful when there are large query sets, not just a few query points.  Instead of
building a tree on the reference set and searching with each query point
separately, Gray and Moore suggest also building a query tree and traversing
both the query and reference trees simultaneously (a {\it dual-tree traversal},
from which the class of algorithms takes its name).

Dual-tree algorithms can be easily understood through the recent framework of
\citet{curtin2013tree}: two trees (a query tree and a reference
tree) are traversed by a {\it pruning dual-tree traversal}.  This traversal
visits combinations of nodes from the trees in some sequence (each combination
consisting of a query node and a reference node), calling a problem-specific
\texttt{Score()} function to determine if the node combination can be pruned.
If not, then a problem-specific \texttt{BaseCase()} function is called for each
combination of points held in the query node and reference node.  This has
significant similarity to the more common single-tree branch-and-bound
algorithms, except that the algroithm must recurse into child nodes of {\it
both} the query tree and reference tree.

There exist numerous dual-tree algorithms for problems as diverse as kernel
density estimation \citep{gray2003nonparametric}, mean shift \citep{wang2007fast},
minimum spanning tree calculation \citep{march2010euclidean}, $n$-point
correlation function estimation \citep{march2012fast}, max-kernel search
\citep{curtin2013fast}, particle smoothing \citep{klaas2006fast}, variational
inference \citep{amizadeh2012variational}, and range search \citep{nbody}, to name
a few.

Some of these algorithms are derived using the cover tree \citep{langford2006}, a
data structure with beneficial theoretical qualities.  Dual-tree all
nearest-neighbor search and approximate kernel density estimation with cover
trees have $O(N)$ runtime guarantees for $O(N)$ queries \citep{ram2009}; minimum
spanning tree calculation scales as $O(N \log N)$ \citep{march2010euclidean}.
Other problems have similar worst-case guarantees \citep{curtin2014dual,
march2013multi}.

In this work we combine the generalization of \citet{curtin2013tree} with the
theoretical results of \citet{langford2006} and others in order to develop a
worst-case runtime bound for any dual-tree algorithm when the cover tree is
used.

Section \ref{sec:preliminaries} lays out the required background, notation, and
introduces the cover tree and its associated theoretical properties.  Readers
familiar with the cover tree literature and dual-tree algorithms
\citep[especially][]{curtin2013tree} may find that section to be review.
Following that, we introduce an intuitive measure of cover tree imbalance in
Section \ref{sec:imbalance}.  This measure of imbalance is then used to prove
the main result of the paper in Section \ref{sec:bound}, which is a worst-case
runtime bound for generalized dual-tree algorithms.  We apply this result to
three specific problems: nearest neighbor search (Section \ref{sec:nns}),
approximate kernel density estimation (Section \ref{sec:akde}), and range search
/ range count (Section \ref{sec:rs}), showing linear runtime bounds for each of
those algorithms.  Each of these bounds is an improvement on the
state-of-the-art, and in the case of range search, is the first such bound.  We
then investigate the empirical performance of cover trees for these algorithms,
with a focus on understanding the dataset-dependent constants that our bounds
use.

@@

%\bibliographystyle{plain}
\bibliography{bounds}
